Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden

### 2020-06-11 output/paper/paper.tex

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency´s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.

@@

\chapter{Model Performance}

Figure~\ref{fig:four_dates} depicts four randomly chosen dates where the model is compared to a benchmark prediction based on historical averages. The model and the benchmark are tasked with predicting the total number of individuals who have died at a given date and have been reported within 14 days of that date. As time progresses, more deaths are reported and the dashed grey line approaches the horizontal line. Meanwhile model uncertainty decreases.

@@

@@

\chapter{Implications and limitations}
The model proposed here has much better ability to estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcome major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing readers how large the range of possible outcomes can be.

@@

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner, given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure compared to Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so. Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days. The number of such cases is very small, however, and its removal should not make much difference. The central assumption of the model is that the proportions deaths reported each day is fixed (up to the known covariates). If actual reporting standards change over time and its not explicitly modelled by a covariate, the model will not be able to account for this. But reporting likely becomes faster as the crisis infrastructure improves. One can imagine that after a while the reporting improves, or is changed, if this is not accounted for by a covariate in the model, it will report incorrect numbers. Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages. When the covariates to the reporting delay pattern is known, the model can incorporate them and provide a more accurate nowcasting of COVID-19 events.

\chapter{Conclusion}

>>>>

In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden. The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average. A version without predictions is used in the Public Health Agency´s daily press briefings. As expected, the model provides estimates of actual deaths considerably above the reported number of deaths. Not how the model predicts additional deaths above the moving average line.

@@

\section{Model Performance}
To say judge whether or not the model is accurate we need to compare it to a benchmark. The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week. Instead, we create a benchmark prediction by adding the mean historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc. We use the sum of standard deviations as confidence intervals, assuming that lags are independent across days. The exact calculation is described in the appendix.}.

Figure~\ref{fig:four_dates} depicts four randomly chosen dates where the model is compared to the benchmark. The model and the benchmark are tasked with predicting the total number of individuals who have died at a given date and have been reported within 14 days of that date. As time progresses, more deaths are reported and the dashed grey line approaches the horizontal line. Meanwhile model uncertainty decreases.

@@

Benchmark and model point estimates are similarly close to the truth. The model produces tighter confidence intervals however, except during the very first day when only a single report is available.

@@

\chapter{Implications and limitations}
The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.

@@

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner. Given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so.

Moreover, deaths are reported from two distinct populations that seem to follow different trends. At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths. But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes.

Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days. The number of such cases is very small, however, and its removal should not make much difference. The central assumption of the model is that the proportions deaths reported each day is fixed (up to the known covariates). If actual reporting standards change over time, the model will not be able to account for this. But reporting likely becomes faster as the crisis infrastructure improves. One can imagine that after a while the reporting improves, or is changed, if this is not accounted for by a covariate in the model, it will report incorrect numbers. Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages. When the covariates to the reporting delay pattern are known, the model can incorporate them and provide more accurate predictions.

\chapter{Conclusion}
In this paper, we provide a simple method to accurately nowcast daily Covid-19 statistics that are reported with delay. By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends. Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.
